---
title: "Exploratory Data Analysis Project"
author: "Nadith Dharmasena"
output: html_notebook
---

### Guiding Question

On average, did the United States of America suffer worse economic consequences than the rest of the world's developed countries? I define "bad" economic consequences as decreased GDP and increased unemployment. Therefore, if the GDP of the US dipped lower and its unemployment rate rose higher on average and percent-wise than other countries, the US would be defined to have suffered "worse" economic consequences. If the reverse is true, then the US would be defined to have enjoyed "better" economic consequences.

### Data Sources Overview

My project examines four data sources to determine the answer to my guiding question. All of the datasets contain economic statistics about OECD countries. OECD countries are generally developed countries, and they collectively comprise the majority of global GDP. Due to their highly developed economies and adept governments, these countries publish detailed economic reports quarter by quarter. The completeness of the economic data made studying these countries more feasible than analyzing developing countries, who rarely publish comprehensive reports.

##### Brief Synopsis of Data Sources

1. GDP - Quarterly nominal GDP growth rate for all OECD countries
2. UNEMPLOYMENT - Ouarterly unemployment rate for all OECD countries
3. MedianListPrice - Median list price of homes in the USA by city
4. Inventory - Inventory of homes in the USA by city
5. COUNTRYCODES - Mapping from country codes to country names

##### Suggested Data Source Analysis Questions

1. Where did you find them?
2. Who collected/maintains them?
3. When & Why were they originally collected?
4. What does a case represent in each data source, and how many total cases are available?
5. What are some of the variables that you plan to use

<i>Note: I did not answer any of the suggested questions for the COUNTRYCODES data set because I only use it to determine a country's name from its country code. I did not use it for any meaningful analysis.</i>

##### GDP

1. OECD iLibrary database
2. The Organization for Economic Cooperation and Development (OECD) collects and maintains the data. It collects this data because each OECD-member country has to submit their economic statistics to the OECD. The OECD simply consolidates all of this submitted data and publishes it on their database.
3. The OECD collected the data from all of the OECD-member countries. Each of the member countries are required to submit their data to the OECD.
4. Each case represents the quarter-to-quarter percentage change of GDP for each quarter since 2018-Q1 for each OECD-member country. Several of the cases also represent quarter-by-quarter GDP percentage change for transnational organizations like the Euroarea, G20, and G7. We can see this in the tables in the "Perusing the Data Tables" section.
5. I plan to use the following variables: LOCATION, INDICATOR, TIME, and Value.

##### UNEMPLOYMENT

1. OECD iLibrary database
2. Like for the GDP data set, the OECD collects, maintains, and publishes the data.
3. Like for the GDP data set, the OECD collects this data from each OECD-member country. 
4. Each case represents the unemployment rate for each quarter since 2018-Q1 for each OECD-member country. Several of the cases also represent quarterly unemployment rate data for transnational organizations like the Euroarea, G20, and G7. We can see this in the tables in the "Perusing the Data Tables" section.
5. I plan to use the following variables: LOCATION, INDICATOR, TIME, and Value.

##### MedianListPrice

1. Zillow Research's website
2. Zillow Research collects, maintains, and publishes the data.
3. Zillow Research gathers this data from home listings on its website. 
4. Each case represents median housing price information for each major US metropolitan area.
5. I plan to use the following variables: RegionName, all of the date columns

##### Inventory

1. Zillow Research's website
2. Zillow Research collects, maintains, and publishes the data.
3. Zillow Research gathers this data from home listings on its website. 
4. Each case represents listed home quantity information for each major US metropolitan area.
5. I plan to use the following variables: RegionName, all of the date columns

### Importing Libraries

```{r}
# clean up workspace environment
rm(list = ls())

# all packages used for the assignment
library(mosaic)
library(tidyverse)
library(zoo)
library(lubridate)
```

### Data Intake

```{r}
GDP <- read.csv("/Users/nadithdharmasena/Desktop/Academics/Junior Year - PSU/First Semester/STAT 184/Project/GDP.csv")
UNEMPL <- read.csv("/Users/nadithdharmasena/Desktop/Academics/Junior Year - PSU/First Semester/STAT 184/Project/UNEMPLOYMENT.csv")
COUNTRYCODES <- read.csv("/Users/nadithdharmasena/Desktop/Academics/Junior Year - PSU/First Semester/STAT 184/Project/COUNTRYCODES.csv")
mlp <- read.csv("/Users/nadithdharmasena/Downloads/MedianListPrice.csv")
inv <- read.csv("/Users/nadithdharmasena/Downloads/Inventory.csv")
```

### Perusing the Data Tables

```{r}
head(GDP)
summary(GDP)

head(UNEMPL)
summary(UNEMPL)

head(mlp)
head(inv)
```

### Cleaning and Augmenting Data Sets

Here, I use the COUNTRYCODES data set to add a country name variable to each case according to its country code. Additionally, I convert some variables from factor type to character string type.

```{r}
GDP_refined <- merge(GDP, COUNTRYCODES, by.x = "LOCATION", by.y = "CODE") %>%
  mutate(
    LOCATION = as.character(LOCATION),
    INDICATOR = as.character(INDICATOR),
    Country = as.character(Country)
  )

UNEMPL_refined <- merge(UNEMPL, COUNTRYCODES, by.x = "LOCATION", by.y = "CODE") %>%
  mutate(
    LOCATION = as.character(LOCATION),
    INDICATOR = as.character(INDICATOR),
    Country = as.character(Country)
  )

GDP_refined

UNEMPL_refined
```


### Determining Average Economic Health

#### United States vs. Other G7 Countries

We can see from the box plot below that the United States has had more consistent GDP growth over the last two decades than most of the other G7 countries. I can make this assertion because the upper 75% of quarterly GDP changes have been positive in the US. Excluding a few outlier quarters, the vast majority of US quarters have had positive GDP growth. Furthermore, the United States has had lower unemployment rates than most other G7 countries over the past two decades. Therefore, it is reasonable to say that the United States economy is one of the strongest in the world in regular times. It also follows that the highly stable economy of the United States should have been able to weather the pandemic much more easily than other countries. 

```{r}
avg_gdp_g7_boxplot <- GDP_refined %>%
  filter(Country %in% c("United States", "France", "Germany", "United Kingdom", "Canada", "Italy", "Japan")) %>%
  filter(Value < 5 & Value > -5) %>%
  ggplot(aes(x = Country, y = Value)) +
  geom_boxplot() +
  theme_bw() +
  ggtitle("G7 Quarterly GDP Growth Rates")

avg_gdp_g7_boxplot

avg_unempl_g7_boxplot <- UNEMPL_refined %>%
  filter(Country %in% c("United States", "France", "Germany", "United Kingdom", "Canada", "Italy", "Japan")) %>%
  ggplot(aes(x = Country, y = Value)) +
  geom_boxplot() +
  theme_bw() +
  ggtitle("G7 Quarterly Unemployment Rates")

avg_unempl_g7_boxplot
```

#### Important Caveat

It is important to note that I only compared the United States to other superpower economies because it does not make sense to compare the US economy to developing ones. Those countries have high growth rates because even small increases in nominal GDP, relative to the major economies' increases, result in large percentage changes. However, their high growth rates do not imply they were as economically stable as the US prior to the coronavirus pandemic. We can see this phenomenon below.

```{r}
avg_gdp_developing_plot <- GDP_refined %>%
  filter(Country %in% c("United States", "Turkey", "Israel", "Brazil", "Argentina", "India")) %>%
  filter(Value < 5 & Value > -5) %>%
  ggplot(aes(x = Country, y = Value)) +
  geom_boxplot() +
  theme_bw() +
  ggtitle("G7 Quarterly GDP Growth Rates")

avg_gdp_developing_plot
```


### 2020-Q2 Economic Health

The 2020-Q2 economic statistics were the first set published following the realization of the coronavirus pandemic. Therefore, this quarter can gauge the impact the pandemic had on any nation's economic health. The two plots and summary statistics below tell a confusing story. On the one hand, the United States outperforms the rest of the OECD, including the G7 countries, for quarterly GDP change. Whereas the US GDP only shrunk by 9%, the rest of the OECD countries' GDPs shrunk by about 10.6%. On the other hand, the United States labor market suffered much more than the rest of the G7 and OECD. The US unemployment rate in 2020 Q2 rose to a little over 12.5%, whereas the OECD had an average unemployment rate of 7.5%. This data demonstrates that although the US economy did not shrink as much as other countries, its labor market suffered far more than the labor markets of other OECD countries. This inconsistency makes it significantly harder to compare the pandemic's economic consequences on the United States to the rest of the developed world.


```{r}
q2_2020_gdpchange.g7plot <- GDP_refined %>%
  filter(Country %in% c("United States", "France", "Germany", "United Kingdom", "Canada", "Italy", "Japan")) %>%
  filter(TIME == "2020-Q2") %>%
  ggplot(aes(x = Country, y = Value)) +
  geom_point() +
  theme_bw() +
  ggtitle("2020-Q2 G7 GDP")

q2_2020_gdpchange.g7plot

q2_2020_gdpchange.excl_us <- GDP_refined %>%
  filter(!(LOCATION %in% c("United States"))) %>%
  filter(TIME == "2020-Q2") %>%
  summarise(avg_change = mean(Value))

q2_2020_gdpchange.excl_us

q2_2020_unempl.g7plot <- UNEMPL_refined %>%
  filter(Country %in% c("United States", "France", "Germany", "United Kingdom", "Canada", "Italy", "Japan")) %>%
  filter(TIME == "2020-Q2") %>%
  ggplot(aes(x = Country, y = Value)) +
  geom_point() +
  theme_bw() +
  ggtitle("2020-Q2 G7 Unemployment")

q2_2020_unempl.g7plot

q2_2020_unempl.excl_us <- UNEMPL_refined %>%
  filter(!(LOCATION %in% c("United States"))) %>%
  filter(TIME == "2020-Q2") %>%
  summarise(average_rate = mean(Value))

q2_2020_unempl.excl_us
```

### Economic Impact in Context

The visualizations below can give the reader insight into how the four largest economies in the world performed in the context of their previous performance. We can see from the first graph, titled "GDP in Context", that the GDP of United States shrunk less drastically than every OECD country except Japan. From the graph titled "Unemployment in Context", we can see that the US unemployment rate far exceeded that of the other mega-economies. 


```{r}
getGDPTimeSeries <- function () {
  
  ret <- GDP_refined %>%
    filter(Country %in% c("United States", "Germany", "United Kingdom", "Japan")) %>%
    mutate(TIME = as.Date(as.yearqtr(TIME, "%Y-Q%q")))
  
  return(ret)
  
}

getUnemplTimeSeries <- function () {
  
  ret <- UNEMPL_refined %>%
    filter(Country %in% c("United States", "Germany", "United Kingdom", "Japan")) %>%
    mutate(TIME = as.Date(as.yearqtr(TIME, "%Y-Q%q")))
  
  return(ret)
  
}
```


```{r}
getGDPTimeSeries() %>%
  ggplot(aes(x = TIME, y = Value, color = Country)) +
  geom_line() +
  xlim(as.Date(as.yearqtr("2018-Q1", "%Y-Q%q")), as.Date(as.yearqtr("2020-Q2", "%Y-Q%q"))) +
  ggtitle("GDP in Context") + 
  theme_bw()


getUnemplTimeSeries() %>%
  ggplot(aes(x = TIME, y = Value, color = Country)) +
  geom_line() +
  xlim(as.Date(as.yearqtr("2018-Q1", "%Y-Q%q")), as.Date(as.yearqtr("2020-Q2", "%Y-Q%q"))) +
  ggtitle("Unemployment in Context") + 
  theme_bw()
```


### Judging Impact on Ordinary People

After noticing the unusually high unemployment rate in the US, I hypothesized that the lack of income among millions of ordinary Americans would exhibit itself in the prices and for-sale quantities of widely-held assets. From general knowledge, I know that ordinary people are heavily invested in homes. Therefore, to judge the impact of the coronavirus pandemic on ordinary people, I chose to analyze home prices and quantity of for-sale homes. Before looking at the data, I reasoned that home prices should have decreased and for-sale home quantity should have increased if the coronavirus pandemic had severely negatively impacted ordinary Americans. This hypothesis resulted from my belief that struggling Americans would have been forced to sell their homes if they lost the ability to pay their mortgages.

The data shows that home prices actually rose in the time following the onset of the pandemic and the quantity of for-sale homes barely changed. Considering the high unemployment rate in the US, this result is surprising. I analyzed the prices and for-sale quantity of homes in the 10 most expensive and 10 least expensive metropolitan markets to gain a broad understanding of how the pandemic affected the US housing market. Clearly, the pandemic did not affected it that much, which leads me to believe that although the unemployment rate was excruciatingly high, the ordinary American was not affected as negatively as would be expected.

```{r}
cols <- colnames(mlp)[6:length(colnames(mlp))]

mlp.refined <- mlp %>%
  select(RegionName, cols) %>%
  gather(key = when, value = MedianPrice, cols) %>% 
  mutate(when = gsub(pattern = "X", replacement = "", when)) %>%
  mutate(Date = lubridate::ymd(when), MedianPrice = MedianPrice / 1000) %>%
  arrange(Date) %>%
  select(RegionName, MedianPrice, Date)

mlp.final <- mlp.refined %>%
  filter(Date == "2020-10-31") %>%
  arrange(desc(MedianPrice))

cheap_cities <- mlp.final$RegionName[87:97]
expensive_cities <- mlp.final$RegionName[1:10]
```

```{r}
showMP <- function (which_cities, title) {
  
  mlp.refined %>%
    filter(RegionName %in% which_cities, Date > lubridate::ymd("2019-06-01")) %>%
    ggplot(aes(x = Date, y = MedianPrice, color = RegionName)) +
    geom_line() +
    theme_bw() +
    ylab("Median Price (K$)") +
    ggtitle(title)
  
}
```

```{r}
showMP(expensive_cities, "Expensive Markets - Median Home Price")
```

```{r}
showMP(cheap_cities, "Cheap Markets - Median Home Price")
```


```{r}
inv.refined <- inv %>%
  select(RegionName, cols) %>%
  gather(key = when, value = Quantity, cols) %>%
  mutate(when = gsub(pattern = "X", replacement = "", when)) %>%
  mutate(Date = lubridate::ymd(when)) %>%
  arrange(Date) %>%
  select(RegionName, Quantity, Date)
```

```{r}
showQuantity <- function (which_cities, title) {
  
  inv.refined %>%
    filter(RegionName %in% which_cities, Date > lubridate::ymd("2019-06-01")) %>%
    ggplot(aes(x = Date, y = Quantity, color = RegionName)) +
    geom_line() +
    theme_bw() +
    ylab("Quantity") +
    ggtitle(title)
  
}
```

```{r}
showQuantity(expensive_cities, "Expensive Markets - Listed Home Quantity")
```


```{r}
showQuantity(cheap_cities, "Cheap Markets - Listed Home Quantity")
```

### Conclusion

Using this data exploration, it appears to me that the United States suffered less than other countries on average from the coronavirus pandemic. I came to this conclusion after observing that the US economy, in terms of nominal GDP, did not shrink as much as the world's other superpower economies. Additionally, the neutral housing market data led me to believe that the ordinary American did not suffer economic consequences as negative as the national unemployment rate suggested. Although the US labor market did suffer significantly more than the other OECD-member countries, it is possible that this problem resulted from the fact that the US decided to go into lock down for a few weeks in Q2. All in all, the US and the American people seem to have fared better than other developed nations during the coronavirus pandemic.





















